\begin{center}
    \textit{Man is condemned to be free; because once thrown into the~world,\\ he is responsible for everything he does} \citep{sartre1956being}.
\end{center}
The names of people, without whom I could not have done this, do not fit on one page. First and foremost, I would like to express my deepest appreciation to you, Audun. It was your calming and guiding presence, which always made me confident that this thing would get to a~successful end, and I wish every student had a~supervisor like you. Ingrid, I cannot convey how thankful I am that you always pushed me towards improvement and never left me alone for it. Thank you, Arnaud, for filling the~role of mentor perfectly, and for making the~time spent in Paris as enjoyable as it was enriching. It was a~great pleasure working with you, Sigrunn, who kept patiently helping, regardless of how many times I have asked the~same questions. Jakub, thank you for your selfless effort, compassion, and for helping me feel at home V Holešovičkách. More people deserve a~mention, besides my supervisors. Special thanks go to Andreas, for your open door and your can-do approach. Many thanks go also to my collaborators, namely Libor, David and Jan, Kristina, Nicole, Amalie, and Alen. Thanks should also go to my office colleagues, the~great folks of Forskningsparken and Bâtiment 14; this would not have been the~same without you. 
\begin{center}
\textit{Having money is a~way of being free of money} \citep{camus1972happy}. 
\end{center}
I would like to express my gratitude to the~funding organizations, and the~project managers, who enabled me to study and pay the~mortgage while doing so. This includes the~funding from Tromsø forskningsstiftelse, Norges forskningsråd, Grantová agentura České republiky, and UiT Norges artiske universitet. I would like to extend my thanks to all the~employees at UiT. You did excellent work at keeping me on track with my study plan. A~great negligence would be not to mention the~open-source community. You are the~giant whose shoulders we all stand on. 
\begin{center}
\textit{Active love is labor and fortitude, and for some people too, perhaps,\\ complete science} \citep{dostoyevskii1880brothers}.
\end{center}
My utmost thankfulness is for my loving family. To my dearest wife Veronika, as you make the~grim days easier, and the~easy days awesome. To my great little sister, Neli, for you never fail to show me a~new perspective. To my grandmother, Ria, for giving me the~foundation stone of independence. Last by certainly not least, to my parents, Adrián and Marína. The~older I grow the~more I see how much you sacrificed in providing me with all I always took for granted. I will not forget that.