This thesis has shown that by statistical analysis, physically meaningful parameters of the~interplanetary dust cloud can be yielded from the~dust impact counts recorded with spacecraft, which improves our understanding of the~inner solar system dust cloud. We showed that a~convolutional neural network is a~viable tool for dust identification in antenna measurements, which we used for SolO data, and which has the~potential to improve the~quality of data products of other spacecraft. We presented a~Bayesian approach to modelling the~dust counts recorded with spacecraft, which can fit more sophisticated models. We also presented and demonstrated a~statistical toolbox for this task, and we used it to characterize the~dynamics of \mbox{$\beta$-meteoroids}, and to describe their deceleration. As a~part of this thesis, the~method of dust detection with antennas was studied, and we have shown the importance of photoelectrons for the~detection. Understanding of the~data relies on robust knowledge of the~impact process, which we also contributed to as a~part of this thesis, by pointing out and the~presence of a two-peak structure in the~impact records. Using the~formalism of kinetic theory, we explained several features of the~dust flux observed with PSP and we showed that such modelling effort is fruitful and potentially practical for other spacecraft and other dust environments. In particular, we showed that the~available data are compatible with a~near-solar dust depletion zone and that the~spacecraft properties are consequential for the~interpretation of the~data. To understand the~inner solar system dust environment more deeply, and to make the~most of the~presented methods, several tasks remain for future inquiry. 

The~dust identification method presented in Paper~I is significantly more reliable than the~methods used before, but it is limited to certain measurement regimes of the~specific device used on SolO. The~development of a~routine for automatic classification of waveform signals is challenging, as the~compatibility between different measurements is limited, due to differences in physical design, and in the~data products. Since human experts can classify waveforms from different devices without technical knowledge of the~device, it is feasible. Having such a~routine would be useful, as it would provide another layer of harmonization of data between different spacecraft. The~application of such routine onboard spacecraft would save the~data transfer, and therefore, potentially allow for better data coverage. 

The~statistical analysis presented in the~thesis and used in Paper~II is superior to the~often used least squares fitting, since it treats the~counting error correctly. This is especially important if the~number of detections within a~temporal interval is a~small number, which it often is. It is strongly suggested that such a~method is used for future analysis of dust counts of not only SolO and PSP. 

The~assumption of a~dust flux proportional to the~impact speed to a~power higher than one was used in several works, including Paper~IV of this thesis. The~reason is that since higher impact speed generates more charge, small grains are detected at a~high impact speed, but not at a~low impact speed. The~flux therefore depends on the~mass distribution of the~grains, and on the~charge production as a~function of the~impact speed. As a~simplification, it was assumed that the~produced charge depends on the~power of the~impact speed, and that the~mass distribution is a~power-law. Both are arguable. The~charge production was never experimentally measured at a~speed as high as is the~typical impact speed on PSP, or even SolO, so we are limited to a~reasonable extrapolation. More questionable is the~assumption of a~power-law distribution of masses. While it might be true for large masses, when grain to grain collisions shape the~cloud, the~dynamics of sub-micron dust depends on the~size in a~different way. This is clearly demonstrated with \mbox{$\beta$-meteoroids}, which move differently from bound dust, but they occupy only an order of magnitude or two on the~mass scale. Since nearly all grains within the~mass range of \mbox{$\beta$-meteoroids} are \mbox{$\beta$-meteoroids}, these are missing in the~mass distribution of bound dust. Therefore, the~mass distribution of micron and sub-micron sized bound dust grains cannot be a~power-law. A~further investigation of their distribution would prove instrumental for modelling efforts. 

Interstellar dust (ISD) was barely addressed in the~thesis, since neither SolO nor PSP are very well suited for its detection. The~situation might change when SolO's orbit becomes more inclined, and gets to the~region, where bound dust and \mbox{$\beta$-meteoroids} are less plentiful. This is coincidentally planned for the~late 2020's, when the~ISD flux will likely have recovered, due to the~orientation of the~solar magnetic field. Understanding the~flux observed out of the~plane of ecliptic would require generalization of the~models for the~flux but will place further constraints on the~dust parameters. We offered tools which are suitable for the~forward modelling of ISD in Paper~IV and for statistical analysis of ISD in Paper~II.

Several spacecraft have reported nanodust observations and since nanodust dynamics are strongly influenced by the~electromagnetic field, its flux is likely to depend on the~solar cycle. We did not observe nanodust with SolO nor with PSP, but this might be due to solar cycle. As was shown in Paper~III, antenna dust detection is specific for each spacecraft, and the~understanding of the~antenna detection process is still limited. However, SolO's electrical suite is similar to that of STEREO, which detected nanodust, so nanodust detection remains an option for SolO in the~future.  

Each spacecraft's in-situ detections happen along its orbit. An intrinsic bond exists between the~velocity and the~location, and, therefore, between the~amount of detected bound dust and \mbox{$\beta$-meteoroids}. Spacecraft, which change their orbital elements due to gravity assists, such as SolO and PSP, change this bond in discrete steps, allowing for a~decoupling the~two components of the~flux from each other. Multi-spacecraft analysis allows for even more, as time and location are not bonded together, allowing for the~estimates of time-evolution of the~dust cloud. As demonstrated in Paper~IV, multi-spacecraft analysis is complicated, but feasible. It is therefore worthy of future pursuit, as SolO will get inclined, and more dust-detecting spacecraft will operate in the~solar system simultaneously. 